Unimodular covers of 3-dimensional parallelepipeds and Cayley sums

### 2019-05-13 unimodular-covers.tex
some edits

\paco{define mixed IDP, Cayley sums, Minkowski sums, mixed subdivisions}
A lattice polytope is called \emph{smooth} if it is simple and the edge generators at every vertex form a linear basis for the lattice. 
%
Oda~\cite{Oda1997} posed several questions regarding IDP polytopes which are now considered conjectures (REFERENCES). 
Among them:
\begin{conjecture}
\label{conj:Oda}
\begin{enumerate}
\item Let $P, Q\subset \R^d$ be lattice polytopes with the normal fan of $Q$ refining that of $P$, and $P$ smooth. Then
\begin{align*}
\label{eq:mixedIDP}
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\end{align*}
\paco{perhaps we can call this formula ``being mixed IDP''. The property of ``$N(Q)$ refines $N(P)$'' is sometimes called ``$P$ is a Minkowski summand of $Q$''.}

@@

\begin{remark}
Higher dimensional parallelotopes (affine images of cubes) are also IDP, but we do not know whether 
they have unimodular covers, or whether $3$-dimensional ones have unimodular triangulations.
\end{remark}

@@

Another case of Oda's questions that has been solved is the two-dimensional case of \Cref{conj:Oda}(1), with three different proofs by
Fakhruddin~\cite{Fakhruddin}, Ogata~\cite{Ogata} and Haase et al.~\cite{HNPS2008}, the latter without assuming $P$ smooth. 
An alternative approach to \Cref{conj:Oda}(1) is via Cayley sums. As we note in \Cref{sec:cayley} (\Cref{prop:mixedIDP}), if $P,Q\subset \R^d$ are two lattice polytopes one has
\[
\cayley(P,Q) \text{ is IDP} 
\Rightarrow
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\]
In \Cref{sec:cayley} we show the following even stronger result, in the two-dimensional case:

@@

In this language, Problems 3 and 4 i \cite{Oda1997} translate to

\begin{question}
If $Q$ is nef for $P$, is the pair $(P,Q)$ mixed IDP?
\end{question}

In dimension two, the answer is always yes by \cite{HNPS2008}. In dimension three it is easy to construct counter-examples (e.g., let $P=Q$ be the empty simplex of volume two), but the case where $P$ is smooth is open, and attributed as a conjecture of Oda.

>>>>

\paco{define mixed IDP, Cayley sums, Minkowski sums, mixed subdivisions}
A lattice polytope is called \emph{smooth} if it is simple and the edge generators at every vertex form a linear basis for the lattice. 
%
Oda~\cite{Oda1997} posed several questions regarding IDP polytopes which are now considered conjectures (REFERENCES). 
Among them:
\begin{conjecture}
\label{conj:Oda}
\begin{enumerate}
\item (Problems 3 and 4 in \cite{Oda1997}) Let $P, Q\subset \R^d$ be lattice polytopes with the normal fan of $Q$ refining that of $P$, and $P$ smooth. Then
\begin{align*}
\label{eq:mixedIDP}
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\end{align*}
\paco{perhaps we can call this formula ``being mixed IDP''. The property of ``$N(Q)$ refines $N(P)$'' is sometimes called ``$P$ is a Minkowski summand of $Q$''.}

@@

\begin{question}
Do $3$-dimensional parallelepipeds have unimodular triangulations?
\end{question}

\begin{question}
Higher dimensional parallelotopes (affine images of cubes) are IDP. Do they have unimodular covers? 
\end{question}

@@

Another case of Oda's questions that has been solved is the two-dimensional case of \Cref{conj:Oda}(1), with three different proofs by
Fakhruddin~\cite{Fakhruddin}, Ogata~\cite{Ogata} and Haase et al.~\cite{HNPS2008}, the latter without assuming $P$ smooth. 
In dimension three however the statement fails without the smoothness assumptin, as shown by letting $P=Q$ be the empty simplex of volume two.

An alternative approach to \Cref{conj:Oda}(1) is via Cayley sums. As we note in \Cref{sec:cayley} (\Cref{prop:mixedIDP}), if $P,Q\subset \R^d$ are two lattice polytopes one has
\[
\cayley(P,Q) \text{ is IDP} 
\Rightarrow
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\]
In \Cref{sec:cayley} we show the following even stronger result, in the two-dimensional case:

@@

In this language, \Cref{conj:Oda}(1) translates to:
\emph{If $Q$ is nef for $P$ and $P$ is smooth then the pair $(P,Q)$ mixed IDP.}
